Adaptive evolution with aneuploidy and mutation

### 2022-06-10 ms/ms.tex
updated truncated prior for kde of hdi and posterior plots

\paragraph{Rates and fitness effects of aneuploidy and mutation.} 
We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets. See \emph{Supporting material} for sensitivity analysis.
%% estimates generated by MAP-and-hdi.ipynb
The estimated mutation rate, $\mu=2.986\cdot10^{-6}\ [3.734\cdot10^{-7}-3.725\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated aneuploidy rate, $\delta=1.724\cdot10^{-3}\ [1.451\cdot10^{-3}-2.766\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).

@@

However, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). %% generated with MAP-and-hdi.ipynb
The estimated mutation rate without aneuploidy is $\mu=7.98\cdot10^{-9}\ [7.91\cdot10^{-9}-8.14\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40. 
The fitness of the mutant is also much lower at $w_{\eumt}=1.013\ [1.012-1.013]$.
This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations. Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).

@@

\paragraph*{Genetic instability in aneuploid cells.}
It has been suggested that aneuploidy increased genetic instability~\citep{Sheltzer2011b}. Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}).
%% generated with MAP-and-hdi.ipynb
With $\tau=100$, the estimated mutation rate was about 7-8-fold lower compared to $\tau=1$ ($\mu=3.84\cdot10^{-7}\ [6.53\cdot10^{-8}-6.207\cdot10^{-7}]$) and the aneuploidy rate was about 2-3-fold lower ($\delta=0.764\cdot10^{-3}\ [0.475\cdot10^{-3}-1.818\cdot10^{-3}]$). 
With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.66\cdot10^{-6}\ [9.033\cdot10^{-8}-2.353\cdot10^{-6}]$). 
WAIC is lowest for $\tau=33/32$ and $\tau=1$.
Therefore, evidence does not support an increase in mutation rate in aneuploid cells, and moreover, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference.
We also checked the differences in genotype frequency dynamics for different $\tau$ values. We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).

@@

% generated with alt-prior.ipynb
We estimated the parameters under this extended informative prior.
Inference took much longer to run but the posterior distribution seemed to converge, as it did not change much in the final iterations. 
The posterior predictive plot shows that inference with this extended prior produces a posterior distribution that fails to explain the empirical observations~(pink in Figure~\ref{fig:2n*-fixation}).
However, the inferred posterior distribution is considerably narrower (compare \Cref{fig:posterior,fig:posterior-alt}) and therefore parameter estimates are less variable.
The estimated mutation rate was much lower compared to the main informative prior, with $\mu=2.475\cdot10^{-9}\ [2.42\cdot10^{-9}-2.609\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.705\cdot10^{-3}\ [2.112\cdot10^{-3}-3.107\cdot10^{-3}]$,
$w_{\anwt}=1.022\ [1.021-1.024]$,
$w_{\anmt}=1.029\ [1.027-1.03]$,
$w_{\eumt}=1.03\ [1.029-1.031]$. 
Notably, the maximum posterior ratio $w_{\eumt}/w_{\anwt}=1.007$ is much lower than the maximum prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in our standard prior.
Together with the posterior predictive results, we conclude that the main informative prior is preferable over the extended informative prior.

>>>>

\paragraph{Rates and fitness effects of aneuploidy and mutation.} 
We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets. See \emph{Supporting material} for sensitivity analysis.
%% estimates generated by MAP-and-hdi.ipynb
The estimated mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).

@@

However, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). %% generated with MAP-and-hdi.ipynb
The estimated mutation rate without aneuploidy is $\mu=7.98\cdot10^{-9}\ [7.906\cdot10^{-9}-8.138\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40. 
The fitness of the mutant is also much lower at $w_{\eumt}=1.013\ [1.012-1.013]$.
This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations. Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).

@@

\paragraph*{Genetic instability in aneuploid cells.}
It has been suggested that aneuploidy increased genetic instability~\citep{Sheltzer2011b}. Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}).
%% generated with MAP-and-hdi.ipynb
With $\tau=100$, the estimated mutation rate was about 7-8-fold lower compared to $\tau=1$ ($\mu=4.094\cdot10^{-7}\ [6.252\cdot10^{-8}-6.046\cdot10^{-7}]$) and the aneuploidy rate was about 2-3-fold lower ($\delta=0.744\cdot10^{-3}\ [0.506\cdot10^{-3}-1.827\cdot10^{-3}]$). 
With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.67\cdot10^{-6}\ [2.836\cdot10^{-8}-2.245\cdot10^{-6}]$). 
WAIC is lowest for $\tau=33/32$ and $\tau=1$.
Therefore, evidence does not support an increase in mutation rate in aneuploid cells, and moreover, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference.
We also checked the differences in genotype frequency dynamics for different $\tau$ values. We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).

@@

% generated with alt-prior.ipynb
We estimated the parameters under this extended informative prior.
Inference took much longer to run but the posterior distribution seemed to converge, as it did not change much in the final iterations. 
The posterior predictive plot shows that inference with this extended prior produces a posterior distribution that fails to explain the empirical observations~(pink in Figure~\ref{fig:2n*-fixation}).
However, the inferred posterior distribution is considerably narrower (compare \Cref{fig:posterior,fig:posterior-alt}) and therefore parameter estimates are less variable.
The estimated mutation rate was much lower compared to the main informative prior, 
% generated with MAP-and-hdi.ipynb
with $\mu=2.475\cdot10^{-9}\ [2.419\cdot10^{-9}-2.609\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.707\cdot10^{-3}\ [2.093\cdot10^{-3}-3.092\cdot10^{-3}]$,
$w_{\anwt}=1.022\ [1.021-1.024]$,
$w_{\anmt}=1.029\ [1.027-1.03]$,
$w_{\eumt}=1.03\ [1.029-1.031]$. 
Notably, the maximum posterior ratio $w_{\eumt}/w_{\anwt}=1.007$ is much lower than the maximum prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in our standard prior.
Together with the posterior predictive results, we conclude that the main informative prior is preferable over the extended informative prior.
